---
title: "Is Costa Rica’s Transition Fast Enough for 1.5°C?"
subtitle: "CO₂, GDP and Transport in a Supposed Green Champion"
author: "Group PG-20"
output:
  xaringan::moon_reader:
    css: ["default", "default-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: "github"
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current% / %total%"
      navigation:
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(scales)
library(ggthemes)
library(knitr)
library(kableExtra)

opts_chunk$set(
	fig.align = "center",
	fig.height = 7,
	fig.width = 14,
	message = FALSE,
	warning = FALSE,
	dev = "svg"
)

theme_set(theme_clean(base_size = 14))
```
---

# Plan

1. **Q1 – Are we on a 1.5°C path?**  
   If we assume a simple linear decline in CO₂ per capita from 2010 to net-zero by 2050, how far is Costa Rica above or below that pathway today?

2. **Q2 – Decoupling compared with similar economies**  
   Among countries that had similar GDP per capita to Costa Rica in 2010, has Costa Rica done unusually well (or not) at growing income while stabilising or reducing CO₂ per capita?

3. **Q3 – Is transport now the main bottleneck?**  
   Given that Costa Rica’s electricity is already mostly low-carbon, to what extent are transport emissions now dominating its remaining CO₂ and limiting further progress?


---

# Data & approach

- **Country focus:** Costa Rica  
- **Core data:** CO₂ emissions, CO₂ per capita, GDP per capita, and sectoral emissions  
- **Source:** Our World in Data – CO₂ and Energy   
- **Period:** From 1960s to early 2020s, with a focus on 2010 onwards for the 1.5°C pathway


**Q1 – 1.5°C pathway**  
Compare observed CO₂ per capita with a **simple linear net-zero-by-2050 path** starting in 2010.


**Q2 & Q3 – Structure and sectors**  
- Build a **peer group** of countries with similar GDP per capita in 2010  
- Compare **GDP vs CO₂ per capita changes** and then zoom in on **transport vs other sectors** for Costa Rica




```{r q1_data, include=FALSE}
# Q1 data + plot object ------------------------------

energy <- read_csv(
  "data/data_raw/owid_energy.csv",
  show_col_types = FALSE
)

co2 <- read_csv(
  "data/data_raw/owid_co2.csv",
  show_col_types = FALSE
)

energy_sel <- energy |>
  dplyr::select(
    country, iso_code, year,
    renewables_share_elec,
    low_carbon_share_elec,
    fossil_share_elec,
    renewables_electricity,  # TWh
    fossil_electricity       # TWh
  )

co2_sel <- co2 |>
  dplyr::select(
    country, iso_code, year,
    co2,                     # total CO2 emissions (Mt)
    co2_per_capita,
    consumption_co2_per_capita,
    gdp,
    population
  )

cr_energy_co2 <- energy_sel |>
  left_join(
    co2_sel,
    by = c("country", "iso_code", "year")) |>
  filter(
    country == "Costa Rica",
    year >= 1990, year <= 2023)

# Get 2010 baseline (if missing, pick nearest year you actually have)
cr_base_2010 <- cr_energy_co2 |>
  filter(year == 2010) |>
  pull(co2_per_capita) |>
  first()

cr_pathway <- cr_energy_co2 |>
  mutate(
    co2pc_target_15c = dplyr::case_when(
      year < 2010 ~ NA_real_,
      year >= 2010 & year <= 2050 ~
        cr_base_2010 * pmax(0, 1 - (year - 2010) / (2050 - 2010)),
      year > 2050 ~ 0
    )
  )

# Trim years to keep only what matters visually
cr_trim <- cr_pathway |>
  filter(year >= 2000)

max_y <- max(cr_trim$co2_per_capita, na.rm = TRUE)

p1 <- ggplot(cr_trim, aes(x = year)) +
  # Light shaded window only for evaluation period
  annotate(
    "rect", xmin = 2010, xmax = 2023,
    ymin = -Inf, ymax = Inf,
    fill = "grey90", alpha = 0.5
  ) +
  # 1.5°C pathway
  geom_line(
    aes(y = co2pc_target_15c),
    colour = "#D55E00",
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  # Observed CO2 per capita
  geom_line(
    aes(y = co2_per_capita),
    colour = "#1A1A1A",
    linewidth = 1.4
  ) +
  # Annotation
  annotate(
    "text",
    x = 2015, y = max_y * 0.55,
    label = "2023 gap: +0.55 t/person\n(~50% above pathway)",
    hjust = 0, size = 4,
    colour = "red"
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    colour = "grey60"
  ) +
  geom_vline(
    xintercept = 2010,
    linetype = "dashed",
    colour = "grey50"
  ) +
  annotate(
    "text", x = 2010, y = 0.15,
    label = "2010 baseline",
    angle = 90, vjust = -0.3, hjust = 0,
    size = 3, colour = "grey40"
  ) +
  labs(
    title = expression(
      "Costa Rica's " * CO[2] * " per capita vs a 1.5" * degree * "C-compatible pathway"
    ),
    subtitle = "Electricity is already low-carbon; reductions now depend on transport and other fossil uses",
    x = "Year",
    y = expression("CO"[2] * " emissions per capita (tonnes/person)"),
    caption = expression("Source: OWID " * CO[2] * " and Energy datasets.")
  ) +
  coord_cartesian(ylim = c(0, max_y * 1.05)) +
  ggthemes::theme_clean(base_size = 14) +
  theme(
    legend.position   = "none",
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(colour = "grey90"),
    plot.title        = element_text(face = "bold"),
    plot.subtitle     = element_text(size = 11, colour = "#555555")
  )

```



---
# Q1 – Are we on a 1.5°C path?

.pull-left[

- I use a **simple stylised pathway**: CO₂ per capita declines **linearly** from 2010 to **net-zero in 2050**.  
- Costa Rica’s **observed CO₂ per capita** is then plotted against this pathway.

- The **gap in the latest year** shows how far Costa Rica is **above** the 1.5°C-style path, under these assumptions.  
- This is not a full carbon budget model – it’s a **transparent, back-of-the-envelope benchmark**.

]


.pull-right[
```{r q1_plot, echo=FALSE, fig.height=6, fig.width=8}
p1
```

]


```{r q2_data, include=FALSE}
# Uses co2_sel from the earlier q1_data chunk

# 2010 GDP per capita for Costa Rica
cr_2010 <- co2_sel |>
  filter(country == "Costa Rica", year == 2010) |>
  filter(!is.na(gdp), !is.na(population)) |>
  mutate(gdp_pc_2010 = gdp / population) |>
  slice(1)

cr_gdp_pc_2010 <- cr_2010$gdp_pc_2010[1]

# Find countries with similar GDP per capita in 2010
peer_candidates <- co2_sel |>
  filter(year == 2010, !is.na(gdp), !is.na(population)) |>
  mutate(gdp_pc_2010 = gdp / population) |>
  mutate(diff = abs(gdp_pc_2010 - cr_gdp_pc_2010)) |>
  arrange(diff)

# Take Costa Rica + 5 closest peers
peer_countries <- peer_candidates |>
  filter(country != "Costa Rica") |>
  slice_head(n = 5) |>
  pull(country)

q2_countries <- c("Costa Rica", peer_countries)

# Panel data, indexing GDP pc and CO2 pc to 2010 = 1
q2_panel <- co2_sel |>
  filter(
    country %in% q2_countries,
    dplyr::between(year, 2010, 2022)
  ) |>
  mutate(
    gdp_pc = gdp / population,
    co2_pc = co2_per_capita
  ) |>
  group_by(country) |>
  mutate(
    base_gdp_pc = gdp_pc[year == 2010][1],
    base_co2_pc = co2_pc[year == 2010][1],
    gdp_pc_index = gdp_pc / base_gdp_pc,
    co2_pc_index = dplyr::if_else(
      base_co2_pc > 0,
      co2_pc / base_co2_pc,
      NA_real_
    )
  ) |>
  ungroup() |>
  pivot_longer(
    cols = c(gdp_pc_index, co2_pc_index),
    names_to = "series",
    values_to = "index"
  ) |>
  mutate(
    series = dplyr::recode(
      series,
      gdp_pc_index = "GDP per capita (2010 = 1)",
      co2_pc_index = "CO\u2082 per capita (2010 = 1)"
    )
  )

# Plot object for Q2
p2 <- ggplot(q2_panel, aes(x = year, y = index, colour = series)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey75") +
  geom_line(linewidth = 1.1) +
  facet_wrap(~ country, ncol = 3) +
  scale_y_continuous(
    breaks = seq(0.5, 2.5, 0.5),
    labels = number_format(accuracy = 0.1)
  ) +
  scale_x_continuous(breaks = c(2010, 2015, 2020, 2022)) +
  labs(
    title    = "GDP vs CO₂ per capita, indexed to 2010 = 1",
    subtitle = "Costa Rica and countries with similar GDP per capita in 2010",
    x        = "Year",
    y        = "Index (2010 = 1)",
    colour   = NULL
  ) +
  theme_clean(base_size = 13) +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(face = "bold")
  )
```

```{r include=FALSE}
decouple_df <- co2_sel |>
  filter(
    country %in% q2_countries,
    year >= 2010,
    !is.na(gdp), !is.na(population),
    !is.na(co2_per_capita)
  ) |>
  mutate(
    gdp_per_capita = gdp / population
  )

summary_decouple <- decouple_df |>
  group_by(country) |>
  arrange(year, .by_group = TRUE) |>
  summarise(
    start_year     = first(year),
    end_year       = last(year),
    gdp_pc_change  = 100 * (last(gdp_per_capita) / first(gdp_per_capita) - 1),
    co2_pc_change  = 100 * (last(co2_per_capita)   /first(co2_per_capita) - 1),
    .groups = "drop"
  ) |>
  arrange(desc(gdp_pc_change))
```




---

# Q2 – GDP vs CO₂ change since 2010

.pull-left[

- Here I summarise **total change from 2010 to the latest year** for each country.  

- **GDP_pc_change**: growth in income per person.  

- **CO₂_pc_change**: change in CO₂ per person.  

- Costa Rica should stand out as **strong GDP growth with flat or slightly falling CO₂**, whereas some peers grow both.

]

.pull-right[

```{r q2_table, echo=FALSE}
summary_decouple |>
  mutate(
    gdp_pc_change = sprintf("%+.1f%%", gdp_pc_change),
    co2_pc_change = sprintf("%+.1f%%", co2_pc_change)
  ) |>
  rename(
    `GDP % change`   = gdp_pc_change,
    `CO2 % change`   = co2_pc_change,
    `Start year`     = start_year,
    `End year`       = end_year
  ) |>
  kable(
    format  = "html",
    align   = c("l", "c", "c", "r", "r"),
    caption = "Change in GDP and CO₂ per capita since 2010"
  ) |>
  kable_styling(full_width = FALSE, font_size = 12) |>
  row_spec(
    which(summary_decouple$country == "Costa Rica"),
    bold = TRUE
  )
```

]


---
## Q2 – Decoupling compared with similar economies

```{r q2_plot_full, echo=FALSE, fig.height=7, fig.width=13}
p2
```








```{r q3_data, include=FALSE}
# Q3 – Transport vs other CO2 for Costa Rica, using OWID grapher transport series

# 1) Transport CO2 series from OWID grapher
transport <- read_csv(
  "https://ourworldindata.org/grapher/co2-emissions-transport.csv",
  show_col_types = FALSE
)

transport_cr <- transport |>
  rename(
    country       = 1,   # "Entity"
    iso_code      = 2,   # "Code"
    year          = 3,   # "Year"
    transport_co2 = 4    # emissions from transport, tonnes
  ) |>
  filter(country == "Costa Rica")

# 2) Costa Rica total CO2 + population from owid_co2 (already read earlier)
cr_transport <- co2 |>
  filter(country == "Costa Rica", year >= 1990) |>
  select(country, iso_code, year, co2, co2_per_capita, population) |>
  left_join(transport_cr, by = c("country", "iso_code", "year")) |>
  mutate(
    # co2 is in *million* tonnes; transport_co2 is in tonnes
    transport_share_total = if_else(
      !is.na(co2) & co2 > 0,
      transport_co2 / (co2 * 1e6),
      NA_real_
    ),
    transport_co2_pc     = transport_co2 / population,
    non_transport_co2_pc = ((co2 * 1e6) - transport_co2) / population
  )

cr_trans_plot <- cr_transport |>
  filter(!is.na(transport_share_total), year >= 1990)

# 3) Plot: transport share of total CO2
p3 <- ggplot(cr_trans_plot, aes(x = year)) +
  geom_line(
    aes(y = transport_share_total * 100),
    colour = "#D55E00",
    linewidth = 1.1
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey70") +
  labs(
    title = "Transport’s share of Costa Rica’s CO₂ emissions",
    x     = "Year",
    y     = "Transport share of total CO₂ (%)"
  ) +
  ggthemes::theme_clean(base_size = 12) +
  theme(
    panel.grid.minor = element_blank()
  )

# 4) Long format for per-capita decomposition
cr_pc_long <- cr_transport |>
  filter(!is.na(transport_co2_pc), year >= 1990) |>
  select(year, transport_co2_pc, non_transport_co2_pc) |>
  pivot_longer(
    cols      = c(transport_co2_pc, non_transport_co2_pc),
    names_to  = "sector",
    values_to = "co2_pc"
  ) |>
  mutate(
    sector = recode(
      sector,
      transport_co2_pc     = "Transport",
      non_transport_co2_pc = "All other sectors"
    )
  )

# 5) Plot: CO2 per capita – transport vs other
p4 <- ggplot(cr_pc_long, aes(x = year, y = co2_pc, colour = sector)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c(
    "Transport"         = "#D55E00",
    "All other sectors" = "#0072B2"
  )) +
  labs(
    title  = "CO₂ per capita from transport vs other sectors",
    x      = "Year",
    y      = "Tonnes CO₂ per person",
    colour = NULL
  ) +
  ggthemes::theme_clean(base_size = 12) +
  theme(
    legend.position   = "bottom",
    panel.grid.minor  = element_blank()
  )

```




---
# Q3 – Has transport become the main bottleneck?

```{r q3_share_plot, echo=FALSE}
p3
```



---
# Q3 – CO₂ per capita: transport vs other sectors

```{r q3_decomp_plot, echo=FALSE}
p4
```









