---
title: "Is Costa Rica’s Transition Fast Enough for 1.5°C?"
subtitle: "CO₂ Trends, GDP Growth, and the Transport Bottleneck"
author: |
  Group PG-20 – MA7419 Coursework
output:
  xaringan::moon_reader:
    self_contained: true
    css: ["default", "default-fonts"]
    nature:
      ratio: "16:9"
      highlightStyle: "github"
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current% / %total%"
      navigation:
---

```{r setup, include=FALSE}
library(tidyverse)
library(plotly)
library(janitor)
library(scales)
library(ggthemes)
library(knitr)
library(kableExtra)
library(ggrepel)

opts_chunk$set(
  fig.align = "center",
  fig.height = 6,
  fig.width = 14,
  message = FALSE,
  warning = FALSE,
  dev = "svg"
)

theme_set(theme_clean(base_size = 14))
```

# Research Questions

**Q1 – Is Costa Rica aligned with a 1.5°C pathway?**  
Compare actual CO₂ per capita to a simple linear 2010–2050 net-zero benchmark.

**Q2 – Has Costa Rica decoupled growth from emissions better than peers?**  
Contrast its GDP and CO₂ per capita changes with countries that had similar GDP per capita in 2010.

**Q3 – What sector is now the binding constraint?**  
Assess whether transport dominates remaining emissions and therefore determines the pace of further reductions.


---

# Data & methodology

**Data sources**  
- Our World in Data CO₂ & Energy datasets (1990–2024)  
- Transport CO₂: OWID “CO₂ emissions from transport” dataset  
- All headline figures expressed per capita unless noted otherwise


**Analytical approach**  
- **Q1:** Construct a stylised linear net-zero path (2010 → 2050) and compare it with Costa Rica’s observed CO₂ per capita  
- **Q2:** Form a peer group with similar 2010 GDP per capita and compare indexed GDP and CO₂ trends  
- **Q3:** Decompose Costa Rica’s emissions into transport vs all other fossil CO₂ to reveal the binding constraint


**Key choices**  
- Use indexed (2010 = 1) values for fair cross-country comparison  
- Focus on per capita metrics to avoid population-size bias  
- Keep the 1.5°C pathway intentionally stylised (illustrative, not prescriptive)


```{r include=FALSE}
# Energy and CO2 datasets

energy <- read_csv(
  "data/data_raw/owid_energy.csv",
  show_col_types = FALSE
)

co2 <- read_csv(
  "data/data_raw/owid_co2.csv",
  show_col_types = FALSE)

energy_sel <- energy |>
  select(
    country, iso_code, year,
    renewables_share_elec,
    low_carbon_share_elec,
    fossil_share_elec,
    renewables_electricity,
    fossil_electricity
  )

co2_sel <- co2 |>
  select(
    country, iso_code, year,
    co2,
    co2_per_capita,
    consumption_co2_per_capita,
    gdp,
    population
  )
```


---
### Global CO₂ Emissions per Capita (1991–2024)
Costa Rica’s per-capita emissions sit in the lower-middle of the global pack – small in totals, but not negligible compared with much larger economies.


```{r chloropleth-world-pc, echo=FALSE}
co2_world <- co2 |>
  filter(
    year >= 1991,
    !is.na(iso_code),
    !is.na(co2),
    !is.na(co2_per_capita),
    population >= 1e6,          # drop tiny states / territories
    co2_per_capita <= 50        # drop extreme outliers
  ) |>
  mutate(
    co2_pc = co2_per_capita,
    hover = if_else(
      country == "Costa Rica",
      paste0(
        "🌟 COSTA RICA 🌟<br>",
        "Year: ", year, "<br>",
        "Total CO₂: ", round(co2, 1), " Mt<br>",
        "CO₂ per capita: ", round(co2_pc, 2), " t"
      ),
      paste0(
        country, "<br>",
        "Year: ", year, "<br>",
        "Total CO₂: ", round(co2, 1), " Mt<br>",
        "CO₂ per capita: ", round(co2_pc, 2), " t"
      )
    )
  )

fig_world_pc <- co2_world |>
  plot_ly(
    type         = "choropleth",
    locations    = ~iso_code,
    locationmode = "ISO-3",
    z            = ~co2_pc,                 
    zmin         = 0,
    zmax         = 25,                     
    frame        = ~year,
    colorscale   = "Coolwarm",
    colorbar     = list(title = "CO₂ per capita (t/person)"),
    text         = ~hover,
    hoverinfo    = "text"
  ) |>
  layout(
    geo = list(
      scope          = "world",
      projection     = list(type = "flat"),
      showframe      = FALSE,
      showcoastlines = TRUE
    )
  )



```

```{r latin-america-map, echo=FALSE, fig.height=6, fig.width=15}
fig_world_pc
```




```{r q1_data, include=FALSE}
# Q1 data + plot object ------------------------------

country_focus <- "Costa Rica"
pathway_start <- 2010
net_zero_year <- 2050
viz_start <- 2000
year_min <- 1990
year_max <- 2023


cr_energy_co2 <- energy_sel |>
  left_join(
    co2_sel,
    by = c("country", "iso_code", "year")) |>
    filter(
    country == country_focus,
    between(year, year_min, year_max))

cr_base <- cr_energy_co2 |>
  filter(year == pathway_start) |>
  pull(co2_per_capita) |>
  first()

cr_pathway <- cr_energy_co2 |>
  mutate(
    co2pc_target_15c = case_when(
      year < pathway_start ~ NA_real_,
      year <= net_zero_year ~
        cr_base * pmax(0, 1 - (year - pathway_start) /
          (net_zero_year - pathway_start)),
      TRUE ~ 0
    )
  )

cr_trim <- cr_pathway |>
  filter(year >= viz_start)

max_y <- max(cr_trim$co2_per_capita, na.rm = TRUE)
latest_year <- max(cr_trim$year, na.rm = TRUE)
latest_values <- cr_trim |>
  filter(year == latest_year) |>
  summarise(
    gap = co2_per_capita - co2pc_target_15c,
    pct = gap / co2pc_target_15c
  )

gap_value <- latest_values$gap[1]
pct_value <- latest_values$pct[1]
pct_label <- if (is.finite(pct_value)) percent(pct_value, accuracy = 1) else "n/a"

gap_label <- sprintf(
  "%d gap: %+0.2f t/person\n(%s above pathway)",
  latest_year,
  gap_value,
  pct_label
)

p1 <- ggplot(cr_trim, aes(x = year)) +
  annotate(
    "rect", xmin = pathway_start, xmax = latest_year,
    ymin = -Inf, ymax = Inf,
    fill = "grey90", alpha = 0.5
  ) +
  geom_line(
    aes(y = co2pc_target_15c),
    colour = "#D55E00",
    linewidth = 1.1,
    linetype = "dashed"
  ) +
  geom_line(
    aes(y = co2_per_capita),
    colour = "#1A1A1A",
    linewidth = 1.4
  ) +
  annotate(
    "text",
    x = pathway_start + 5, y = max_y * 0.55,
    label = gap_label,
    hjust = 0, size = 4,
    colour = "red"
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    colour = "grey60"
  ) +
  geom_vline(
    xintercept = pathway_start,
    linetype = "dashed",
    colour = "grey50"
  ) +
  annotate(
    "text", x = pathway_start, y = 0.15,
    label = "2010 baseline",
    angle = 90, vjust = -0.3, hjust = 0,
    size = 3, colour = "grey40"
  ) +
  labs(
    title = "Costa Rica's CO₂ per capita vs a 1.5C -compatible pathway",
    subtitle = "Electricity is already low-carbon; reductions now depend on transport and other fossil uses",
    x = "Year",
    y = "CO₂ emissions per capita (tonnes/person)",
    caption = "Source: OWID CO₂ and Energy datasets.") +
  coord_cartesian(ylim = c(0, max_y * 1.05)) +
  theme_clean(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour = "grey90"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(size = 11, colour = "#555555")
  )

```



---
# Q1 – Is Costa Rica on a 1.5°C path?
Costa Rica’s 2023 CO₂ per capita is ~50% above the stylised 1.5°C-compatible pathway, implying deeper cuts are now required.

```{r q1_plot, echo=FALSE, fig.height=6, fig.width=14}
ggplotly(p1)
```



```{r q2_data, include=FALSE}
# Uses co2_sel from the earlier q1_data chunk

# 2010 GDP per capita for Costa Rica
cr_2010 <- co2_sel |>
  filter(country == "Costa Rica", year == 2010) |>
  filter(!is.na(gdp), !is.na(population)) |>
  mutate(gdp_pc_2010 = gdp / population) |>
  slice(1)

cr_gdp_pc_2010 <- cr_2010$gdp_pc_2010[1]

# Candidate pool with GDP differences
peer_candidates <- co2_sel |>
  filter(year == 2010, !is.na(gdp), !is.na(population)) |>
  mutate(
    gdp_pc_2010 = gdp / population,
    diff = abs(gdp_pc_2010 - cr_gdp_pc_2010)
  )

# Scatter peers: all countries within ±20% band
peer_band <- 0.20
peer_band_countries <- peer_candidates |>
  filter(
    country != "Costa Rica",
    between(
      gdp_pc_2010,
      cr_gdp_pc_2010 * (1 - peer_band),
      cr_gdp_pc_2010 * (1 + peer_band)
    )
  ) |>
  arrange(diff) |>
  pull(country)

# Line-chart peers: closest five countries by GDP per capita
peer_top5 <- peer_candidates |>
  filter(country != "Costa Rica") |>
  arrange(diff) |>
  slice_head(n = 5) |>
  pull(country)

q2_countries_scatter <- unique(c("Costa Rica", peer_band_countries))
q2_countries_line <- unique(c("Costa Rica", peer_top5))

# Panel data for line plot peers, indexing GDP pc and CO2 pc to 2010 = 1
q2_panel_line <- co2_sel |>
  filter(
    country %in% q2_countries_line,
    between(year, 2010, 2024)
  ) |>
  mutate(
    gdp_pc = gdp / population,
    co2_pc = co2_per_capita
  ) |>
  group_by(country) |>
  mutate(
    base_gdp_pc = gdp_pc[year == 2010][1],
    base_co2_pc = co2_pc[year == 2010][1],
    gdp_pc_index = gdp_pc / base_gdp_pc,
    co2_pc_index = if_else(
      base_co2_pc > 0,
      co2_pc / base_co2_pc,
      NA_real_
    )
  ) |>
  ungroup() |>
  pivot_longer(
    cols = c(gdp_pc_index, co2_pc_index),
    names_to = "series",
    values_to = "index"
  ) |>
  mutate(
    series = dplyr::recode(
      series,
      gdp_pc_index = "GDP per capita (2010 = 1)",
      co2_pc_index = "CO\u2082 per capita (2010 = 1)"
    )
  )

# Add presentation variables for hover text and styling
q2_panel_line <- q2_panel_line |>
  mutate(
    highlight = country == "Costa Rica",
    hover_text = paste0(
      "Country: ", country,
      "<br>", series,
      "<br>Year: ", year,
      "<br>Index: ", scales::number(index, accuracy = 0.01)
    )
  )
```

```{r include=FALSE}
decouple_df <- co2_sel |>
  filter(
    country %in% q2_countries_scatter,
    year >= 2010,
    !is.na(gdp), !is.na(population),
    !is.na(co2_per_capita)
  ) |>
  mutate(
    gdp_per_capita = gdp / population
  )

summary_decouple <- decouple_df |>
  group_by(country) |>
  arrange(year, .by_group = TRUE) |>
  summarise(
    start_year     = first(year),
    end_year       = last(year),
    gdp_pc_change  = 100 * (last(gdp_per_capita) / first(gdp_per_capita) - 1),
    co2_pc_change  = 100 * (last(co2_per_capita)   /first(co2_per_capita) - 1),
    .groups = "drop"
  ) |>
  arrange(desc(gdp_pc_change))
```

```{r q2_scatter_data, include=FALSE}
q2_scatter <- summary_decouple |>
  mutate(
    highlight = country == "Costa Rica",
    hover_text = sprintf(
      "Country: %s<br>GDP change: %+.1f%%%%<br>CO\u2082 change: %+.1f%%%%",
      country,
      gdp_pc_change,
      co2_pc_change
    ),
    label = country
  )

scatter_peers <- q2_scatter |> filter(!highlight)
scatter_cr <- q2_scatter |> filter(highlight)

scatter_x_range <- range(q2_scatter$gdp_pc_change, na.rm = TRUE)
scatter_y_range <- range(q2_scatter$co2_pc_change, na.rm = TRUE)

p2_scatter <- plot_ly()

if (nrow(scatter_peers) > 0) {
  p2_scatter <- p2_scatter |>
    add_trace(
      data = scatter_peers,
      x = ~gdp_pc_change,
      y = ~co2_pc_change,
      type = "scatter",
      mode = "markers+text",
      text = ~country,
      textposition = "top center",
      textfont = list(color = "#666666", size = 11),
      hovertext = ~hover_text,
      hoverinfo = "text",
      marker = list(
        color = "#AEB6C4",
        size = 10,
        opacity = 0.85
      ),
      showlegend = FALSE
    )
}

if (nrow(scatter_cr) > 0) {
  p2_scatter <- p2_scatter |>
    add_trace(
      data = scatter_cr,
      x = ~gdp_pc_change,
      y = ~co2_pc_change,
      type = "scatter",
      mode = "markers+text",
      text = ~country,
      textposition = "top center",
      textfont = list(color = "#D55E00", size = 12, family = "Arial Black"),
      hovertext = ~hover_text,
      hoverinfo = "text",
      marker = list(
        color = "#D55E00",
        size = 12
      ),
      name = "Costa Rica",
      showlegend = FALSE
    )
}

p2_scatter <- p2_scatter |>
  layout(
    title = "GDP vs CO₂ per capita change since 2010",
    xaxis = list(
      title = "GDP per capita change (%)",
      zeroline = TRUE,
      zerolinecolor = "#CCCCCC"
    ),
    yaxis = list(
      title = "CO₂ per capita change (%)",
      zeroline = TRUE,
      zerolinecolor = "#CCCCCC"
    ),
    shapes = list(
      list(
        type = "line",
        x0 = 0, x1 = 0,
        y0 = scatter_y_range[1], y1 = scatter_y_range[2],
        line = list(color = "#B0B0B0", dash = "dot")
      ),
      list(
        type = "line",
        y0 = 0, y1 = 0,
        x0 = scatter_x_range[1], x1 = scatter_x_range[2],
        line = list(color = "#B0B0B0", dash = "dot")
      )
    )
  )

build_series_plot <- function(df_series, y_label, show_cr_legend = FALSE) {
  peers <- df_series |> filter(country != "Costa Rica")
  costa <- df_series |> filter(country == "Costa Rica")

  plot <- plot_ly()

  if (nrow(peers) > 0) {
    for (ctry in unique(peers$country)) {
      ctry_df <- peers |> filter(country == ctry)
      plot <- plot |>
        add_trace(
          data = ctry_df,
          x = ~year,
          y = ~index,
          type = "scatter",
          mode = "lines",
          name = ctry,
          legendgroup = ctry,
          line = list(color = "#B3BBCB", width = 2),
          opacity = 0.45,
          hovertext = ~hover_text,
          hoverinfo = "text",
          showlegend = FALSE
        )
    }
  }

  if (nrow(costa) > 0) {
    plot <- plot |>
      add_trace(
        data = costa,
        x = ~year,
        y = ~index,
        type = "scatter",
        mode = "lines",
        name = "Costa Rica",
        legendgroup = "Costa Rica",
        line = list(color = "#D55E00", width = 4),
        hovertext = ~hover_text,
        hoverinfo = "text",
        showlegend = show_cr_legend
      )
  }

  plot |>
    layout(
      yaxis = list(title = y_label, zeroline = FALSE),
      xaxis = list(title = "Year")
    )
}

gdp_series <- q2_panel_line |> filter(series == "GDP per capita (2010 = 1)")
co2_series <- q2_panel_line |> filter(series == "CO\u2082 per capita (2010 = 1)")

p2_line <- subplot(
  build_series_plot(gdp_series, "GDP per capita (2010 = 1)", show_cr_legend = TRUE),
  build_series_plot(co2_series, "CO\u2082 per capita (2010 = 1)"),
  nrows = 2,
  shareX = TRUE,
  titleX = TRUE,
  titleY = TRUE
) |>
  layout(
    title = "Indexed GDP and CO₂ per capita (2010 = 1)",
    margin = list(t = 60)
  )
```




---

# Q2 – GDP vs CO₂ change since 2010
Costa Rica delivers ~32% GDP per capita growth while keeping CO₂ per capita just below 2010 levels—better than many similar-income peers but not unique.

```{r q2_scatter_plot, echo=FALSE, fig.height=6, fig.width=14}
p2_scatter
```











```{r q3_data, include=FALSE}
# Q3 – Transport vs other CO2 for Costa Rica, using OWID grapher transport series

# 1) Transport CO2 series from OWID grapher
transport <- read_csv(
  "https://ourworldindata.org/grapher/co2-emissions-transport.csv",
  show_col_types = FALSE
)

transport_cr <- transport |>
  rename(
    country       = 1,   # "Entity"
    iso_code      = 2,   # "Code"
    year          = 3,   # "Year"
    transport_co2 = 4    # emissions from transport, tonnes
  ) |>
  filter(country == "Costa Rica")

# 2) Costa Rica total CO2 + population from owid_co2 (already read earlier)
cr_transport <- co2 |>
  filter(country == "Costa Rica", year >= 1990) |>
  select(country, iso_code, year, co2, co2_per_capita, population) |>
  left_join(transport_cr, by = c("country", "iso_code", "year")) |>
  mutate(
    # co2 is in *million* tonnes; transport_co2 is in tonnes
    transport_share_total = if_else(
      !is.na(co2) & co2 > 0,
      transport_co2 / (co2 * 1e6),
      NA_real_
    ),
    transport_co2_pc     = transport_co2 / population,
    non_transport_co2_pc = ((co2 * 1e6) - transport_co2) / population
  )

cr_trans_plot <- cr_transport |>
  filter(!is.na(transport_share_total), year >= 1990)

# 3) Plot: transport share of total CO2
p3 <- ggplot(cr_trans_plot, aes(x = year)) +
  geom_line(
    aes(y = transport_share_total * 100),
    colour = "#D55E00",
    linewidth = 1.1
  ) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey70") +
  labs(
    title = "Transport’s share of Costa Rica’s CO₂ emissions",
    x     = "Year",
    y     = "Transport share of total CO₂ (%)"
  ) +
  ggthemes::theme_clean(base_size = 12) +
  theme(
    panel.grid.minor = element_blank()
  )

# 4) Long format for per-capita decomposition
cr_pc_long <- cr_transport |>
  filter(!is.na(transport_co2_pc), year >= 1990) |>
  select(year, transport_co2_pc, non_transport_co2_pc) |>
  pivot_longer(
    cols      = c(transport_co2_pc, non_transport_co2_pc),
    names_to  = "sector",
    values_to = "co2_pc"
  ) |>
  mutate(
    sector = recode(
      sector,
      transport_co2_pc     = "Transport",
      non_transport_co2_pc = "All other sectors"
    )
  )

# 5) Plot: CO2 per capita – transport vs other
p4 <- ggplot(cr_pc_long, aes(x = year, y = co2_pc, colour = sector)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c(
    "Transport"         = "#D55E00",
    "All other sectors" = "#0072B2"
  )) +
  labs(
    title  = "CO₂ per capita from transport vs other sectors",
    x      = "Year",
    y      = "Tonnes CO₂ per person",
    colour = NULL
  ) +
  ggthemes::theme_clean(base_size = 12) +
  theme(
    legend.position   = "bottom",
    panel.grid.minor  = element_blank()
  )

```




---
# Q3 – Has transport become the main bottleneck?
By 2020s, ~70–75% of Costa Rica’s fossil CO₂ comes from transport – almost everything left is on the road.
```{r q3_share_plot, echo=FALSE}
ggplotly(p3)
```

---
# Q3 – CO₂ per capita: transport vs other sectors

Other sectors decline or flatten; transport hardly falls, making it the main brake on further progress
```{r q3_decomp_plot, echo=FALSE}
ggplotly(p4)
```


---
### Discussion

- **Q1 – 1.5°C pathway**  
  2023 CO₂ per capita is ~50% above a simple net-zero-by-2050 path from 2010 → future cuts must be steeper than in the past decade.

- **Q2 – Relative decoupling**  
  Versus similar-income economies in 2010, Costa Rica combines ~32% GDP per-capita growth with flat/slightly falling CO₂ per capita – better than many peers, but not unique.

- **Q3 – Transport as bottleneck**  
  Transport now accounts for ~70–75% of Costa Rica’s CO₂; other sectors are flat or falling, so further progress depends mainly on transport.

- **Limitations**  
  Stylised per-capita pathway; sensitive to baseline year, peer group and production- vs consumption-based emissions, but the overall story is robust.

---

### Conclusion – Is Costa Rica’s transition fast enough for 1.5°C?

- **Overall verdict:** Under our simple 1.5°C-style path, no – Costa Rica is overshooting its implied per-capita emissions level by roughly half, so continuing at the recent pace is not fast enough.

- **Relative performance:** Among similar-income economies since 2010, Costa Rica looks relatively well-behaved – richer without higher CO₂ per person – but still not aligned with rapid global decarbonisation.

- **What must change:** With electricity mostly low-carbon, further progress depends on transport: faster vehicle electrification, cleaner fuels and modal shift. Without that, Costa Rica remains a “green champion” more in branding than in 1.5°C-compatible trajectories.





